---
title: "EDA of Monthly milk production dataset"
output: html_notebook
---
```{r Import libraries, warning=FALSE}
# importing libraries

library(tsdl)
library(forecast)
library(dygraphs)
```

The dataset is present in the tsdl library of R.

```{r Call Dataset}
d = subset(tsdl, description="Monthly milk production")

# Details of the dataset
print(attributes(d[[1]])$description)
```

A look at the form of the dataset 

```{r}
d[[1]]
```

A basic plot is first drawn to have an idea of the structure of the time series.

```{r Plot 1}
autoplot(d[[1]])
```
The plot displays both trend and seasonality. The seasonal trend seems similar throughout, increasing with each year.

The data is clean and does not need any work. It is also of the time series "ts" class, and hence does not need any transformation.

```{r}
class(d[[1]])
```
```{r Plot 2}
plot(d[[1]])
abline(reg=lm(d[[1]]~time(d[[1]]))) # fit a trend line
```

Fitting a linear curve through the plot shows an increasing trend, with no outliers or unexpected points of concern.

```{r}
cycle(d[[1]])
```

Here, it is noted that the order of months in determining the seasonality of the data is same.If we consider the 12 months to form a cycle,then the months of 1962 are  ordered in the sequence they appear. With each subsequent year, the months are categorized as to which month of 1962 they correspond to.For eg., Jan 1963 is similar to the value of Jan 1962, and hence, takes its value in the cycle. 

There is no irregularity in the pattern of seasonality throughout the years.This conclusion is strengthened in the plot below.

```{r Plot 3}
sa = seasadj(ts_decom)
seasonplot(sa, 12, col=rainbow(12), year.labels=TRUE, main="Seasonal plot: Airpassengers") # seasonal frequency set as 12 for monthly data.
```

The above plot shows the data after removing the seasonal component. It is to be observed that the points are linear in nature, with the line moving upwards(due to increase in value) with each passing year.

The observations wrt trend and seasonality are fortified when the time series data is decomposed into its components as per the additive model.

```{r Plot 4}
ts_decom = decompose(d[[1]])
plot(ts_decom)
```

The irregularity is random in nature and hence, not a point of concern.

A boxplot is made to have a deeper look into the seasonality of each month.

```{r Plot 5}
boxplot(d[[1]]~cycle(d[[1]]))
```

The general shape throughout the year turns out to be very similar to that observed for seasonality in plot 4 corresponding to each year. 




